Determinism, Complexity, and Predictability in Computer Performance

### 2014-02-26 intro.tex
Adding to and editing intro

@@

Computers are among the most complex engineered artifacts in current
use.  Modern microprocessor chips contain multiple processing units
and multi-layer memories, for instance, and they use complicated
hardware/software strategies to move data and threads of computation
across those resources.  These features---along with all the others
that go into the design of these chips---make the patterns of their
processor loads and memory accesses highly complex and hard to
predict.  Accurate forecasts of these quantities, if one could
construct them, could be used to improve computer design.  If one
could predict that a particular computational thread would be bogged
down for the next 0.6 seconds waiting for data from main memory, for
instance, one could save power by putting that thread on hold for that
time period (e.g., by migrating it to a processing unit whose clock
speed is scaled back).  Computer performance traces are, however, very
complex.  Even a simple ``microkernel,'' like a three-line loop that
repeatedly initializes a matrix in column-major order, can produce
{\sl chaotic} performance traces \cite{mytkowicz09}, as shown in
Figure~\ref{fig:col-ts}, and chaos places fundamental limits on
predictability.
%
 \begin{figure}[htbp]
    \centering
    \includegraphics[width=\textwidth]{figs/colshortts}
    % where an .eps filename suffix will be assumed under latex,
    % and a .pdf suffix will be assumed for pdflatex
    \caption{A small snippet of the instructions per cycle(ipc) of {\tt
        \col}, a three-line C program that repeatedly initializes
      a matrix in column-major order, running on an Intel i7\textsuperscript{\textregistered}-based machine.  Even this
      simple program exhibits chaotic performance dynamics.}
   \label{fig:ipc}
  \end{figure}

The computer systems community has applied a variety of prediction
strategies to traces like this, most of which employ regression.  An
appealing alternative builds on the recently established fact that
computers can be effectively modeled as deterministic nonlinear
dynamical systems \cite{mytkowicz09}.  This result implies the
existence of a deterministic forecast rule for those dynamics.  In
particular, one can use \emph{delay-coordinate embedding} to
reconstruct the underlying dynamics of computer performance, then use
the resulting model to forecast the future values of computer
performance metrics such as memory or processor loads
\cite{josh-ida2011}.  In the case of simple microkernels like the one
that produced the trace in Figure~\ref{fig:ipc}, this deterministic
modeling and forecast strategy works very well.  In more-complicated
programs, however, such as speech recognition software or compilers,
this forecast strategy---as well as the traditional methods---break
down quickly.

This paper is a first step in understanding when, why, and how
deterministic forecast strategies fail when they are applied to
deterministic systems.  We focus here on the specific example of
computer performance.  We conjecture that the complexity of traces
from these systems---which results from the inherent dimension,
non-linearity, and non-stationarity of the dynamics, as well as from
measurement issues like noise, aggregation, and finite data
length---can make those deterministic signals \emph{effectively}
unpredictable.  We argue that \emph{permutation entropy}
\cite{bandt2002per}, a method for measuring the entropy of a
real-valued-finite-length time series through ordinal analysis, is an
effective way to explore that conjecture.  We study four
examples---two simple microkernels and two complex programs from the
SPEC benchmark suite---running on different Intel-based machines.  For
each program, we calculate the permutation entropy of the processor
load (instructions per cycle) and memory-use efficiency (cache-miss
rates), then compare that to the prediction accuracy attainable for
that trace using a simple deterministic model.

@@

% The rest of the paper is organized as follows.
% Section~\ref{sec:compModel} describes the experimental setup, as well
% as the nonlinear modeling and forecast strategies.  In
% Section~\ref{sec:meaComplex}, we review permutation entropy, calculate
% its value for a number of different computer performance traces, and
% compare the results to the prediction accuracy.  In
% Section~\ref{sec:conc}, we discuss these results and their
% implications in regard to our conjecture, and consider future areas of
% research.

>>>>

\end{enumerate}
\noindent ***NOTES END HERE****

\noindent ****SECTION START****

Complicated time series are ubiquitous in modern scientific research. 
%Characterizing and forecasting the future behavior of these observed systems is a vastly researched area with many open questions. 
Every observed time series, complicated or not, exists on a spectrum of \emph{complexity}\footnote{An approx of (blah ) entropy which we make more rigorous in blah}. On the low end of that spectrum are time series that exhibit perfect predictive structure, i.e, signals whose future value are perfectly predicted by knowledge of past values from some ``ideal" model. In particular, there is a an underlying process that generates and transmits information from the past to the future in a perfectly predictable fashion. Some examples of this are constant or low-period signals. On the opposite end of this spectrum are signals which are ``fully complex", i.e., the underlying generating process creates information so rapidly that no information at all is transmitted from the past to the future. Some examples, of this are white noise or random walk processes. With signals on this side of the spectrum, knowledge of the past gives absolutely no insight into the future, \emph{regardless} of the chosen model.  In the middle of this spectrum are where complicated \emph{and} forecast-able signals exist, e.g., deterministic chaotic trajectories, high period signals with some level of noise on top. With time series in this portion of the spectrum, enough information is being transmitted from the past to the future that given the \emph{ideal} model the future of the observed system could be forecast with high accuracy.

Unfortunately, as a corollary of the undecidability of the halting problem: there cannot exist \emph{one} universally ideal forecasting scheme for even completely noise-free deterministic time series\cite{sfi-prediction}, let alone an arbitrary time series. This naturally leads to an important and hard question: given a complicated real-valued time series, is it possible to reliably quantify where on this spectrum the time series lies? i.e., when little or nothing is known about the underlying system's generating process (e.g., linear, nonlinear, deterministic, stochastic, stationary, non-stationary etc.) how can we infer from the time series itself if the signal is too \emph{complex} to forecast?

To answer this question and for the purposes of this paper we focus on real-valued (possibly) noisy scalar time series which appear \emph{complicated}, i.e., they exhibit interesting behavior, for instance, not of low period, strictly monotonic, or constant. We make no assumptions about the generating process's properties,(e.g., linear, nonlinear, deterministic, stochastic, stationary, non-stationary, etc.) and we attempt to measure empirically where on the previously mentioned spectrum of \emph{complexity} the time series exists. For the remainder of this paper we define complexity as a particular approximation of Shannon \emph{entropy} [[Ryan is this right or is it Renyi?]], which we define and explain rigorously in Section \ref{sec:meaComplex}.  To validate these claims we model the time series using a variety of forecasting models which we introduce in Section \ref{sec:model}. We then compare the empirically estimated \emph{complexity} with the accuracy of each forecast method. Which results in two primary findings:

\begin{enumerate}

\item The complexity of a noisy real-valued time series is quantifiable.

\item The way information is processed internally by a given process plays a crucial role in the success of different forecasting schema.

@@

%[[Introduce and make it clear why we choose comp. perf. as test bed.]]

The ideal test bed for this purpose are computer performance traces. Computers are among the most complex engineered artifacts in current
use.  Modern microprocessor chips contain multiple processing units
and multi-layer memories, for instance, and they use complicated
hardware/software strategies to move data and threads of computation
across those resources.  These features---along with all the others
that go into the design of these chips---make the patterns of their
processor loads highly complicated. Accurate forecasts of these quantities, if one could
construct them, could be used to improve computer design. 
% If one
%could predict that a particular computational thread %would be bogged
%down for the next 0.6 seconds waiting for data from main memory, for
%instance, one could save power by putting that thread %on hold for that
%time period (e.g., by migrating it to a processing %unit whose clock
%speed is scaled back). 
 Computer performance traces are, however, very
complicated and range across the entire spectrum of complexity, making this an ideal test bed.  Even a simple ``microkernel,'' like a three-line loop that
repeatedly initializes a matrix in column-major order, can produce everything from periodic to
{\sl chaotic} performance traces \cite{mytkowicz09}, depending on the architecture running the software. Such a performance trace is shown in
Figure~\ref{fig:ipc}.%, and chaos places fundamental limits on predictability.
%
 \begin{figure}[htbp]
    \centering
    \includegraphics[width=\textwidth]{figs/colshortts}
    % where an .eps filename suffix will be assumed under latex,
    % and a .pdf suffix will be assumed for pdflatex
    \caption{A small snippet of the instructions per cycle(ipc) of {\tt
        \col}, a three-line C program that repeatedly initializes
      a matrix in column-major order, running on an Intel i7\textsuperscript{\textregistered}-based machine.  Even this
      simple program exhibits chaotic performance dynamics.}
   \label{fig:ipc}
  \end{figure}
With a single system producing complicated time series that vary so greatly in complexity, it would be invaluable to known a priori if a trace contained enough predictive structure to forecast before time and effort was spent on trying to model and forecast it.

The computer systems community has applied a variety of prediction
strategies to traces like this, most of which employ regression.  An
appealing alternative builds on the recently established fact that
computers can be effectively modeled as deterministic nonlinear
dynamical systems \cite{mytkowicz09}.  This result implies the
existence of a deterministic forecast rule for those dynamics.  In
particular, one can use \emph{delay-coordinate embedding} to
reconstruct the underlying dynamics of computer performance, then use
the resulting model to forecast the future values of computer
performance metrics such as memory or processor loads
\cite{josh-ida2011}.  In the case of simple microkernels like the one
that produced the trace in Figure~\ref{fig:ipc}, this deterministic
modeling and forecast strategy works very well.  In more-complicated
programs, however, such as numerical software or compilers,
this forecast strategy---as well as the traditional methods---break
down some of the time, but work fine others.

This paper is a first step in understanding when, why, and how
deterministic forecast strategies fail when they are applied to
deterministic systems.  We focus here on the specific example of
computer performance but believe the results apply to a much broader class of time series.  We conjecture that the complexity of traces
from these systems---which results from the inherent dimension,
non-linearity, and non-stationarity of the dynamics, as well as from
measurement issues like noise, aggregation, and finite data
length---can make those deterministic signals \emph{effectively}
unpredictable.  We argue that \emph{permutation entropy}
\cite{bandt2002per}, a method for measuring the entropy of a
real-valued-finite-length time series through ordinal analysis, is an
effective way to explore that conjecture.  We study three
examples---a simple microkernel and two complex programs: one from the
SPEC 2006CPU benchmark suite, and one from LAPACK---running on an Intel i7-based machine.  For
each program, we calculate the permutation entropy of the processor
load (instructions per cycle), then compare that to the prediction accuracy attainable for
that trace using a series of deterministic models.

@@

 The rest of the paper is organized as follows.
 Section~\ref{sec:compModel} describes the experimental setup, as well
 as the nonlinear modeling and forecast strategies.  In
 Section~\ref{sec:meaComplex}, we review permutation entropy, calculate
 its value for a number of different computer performance traces, and
 compare the results to the prediction accuracy.  In
 Section~\ref{sec:conc}, we discuss these results and their
 implications in regard to our conjecture, and consider future areas of
 research.
